---
title: "Bill and Flipper Morphology of Palmer Archipelago Penguins"
subtitle: "A Quarto report exploring body size relationships across three penguin species"
author: "Yang An"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-location: left
    code-fold: true
    code-summary: "Show the code"
    fig-align: center
    embed-resources: true
knitr:
  opts_chunk:
    fig.path: "../output/"
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

# Load all packages needed for this report.
# - tidyverse: data wrangling (dplyr) and plotting (ggplot2)
# - here: build file paths relative to the project root, not the working directory
# - palmerpenguins: the dataset used throughout this report
# - kableExtra: turn a plain data frame into a nicely styled HTML table
library(tidyverse)
library(here)
library(palmerpenguins)
library(kableExtra)
```

## Introduction

This report revisits the **`penguins`** dataset from the `palmerpenguins`
package, which has already been used earlier in this course to explore
flipper length and body mass. Here the focus shifts to a relationship we
have *not* yet looked at closely: how **bill length** and **flipper
length** relate to one another across the three species recorded at the
Palmer Long-Term Ecological Research station in Antarctica -- *Adelie*,
*Chinstrap*, and *Gentoo* penguins.

The goals of this report are to:

- clean the data and remove any incomplete records,
- summarize body size by species and sex in a clean, readable table, and
- visualize the relationship between bill length and flipper length with
  a publication-quality figure.

## Data Preparation

Before summarizing or plotting anything, the raw `penguins` data are
cleaned. Rows with missing measurements or unrecorded sex are dropped so
that every summary statistic below is calculated from **complete cases
only**. This keeps the table and the figure consistent with one another.

```{r}
#| label: clean-data

# Drop any row with a missing value in the columns we actually use.
# This avoids NA-driven warnings later and keeps the table and the
# figure based on an identical set of individuals.
penguins_clean <- penguins |>
  drop_na(bill_length_mm, flipper_length_mm, body_mass_g, sex) |>
  mutate(
    species = as.factor(species),
    sex = as.factor(sex)
  )
```

After cleaning, the dataset contains `r nrow(penguins_clean)` penguins
(out of `r nrow(penguins)` originally recorded), spread across three
species and two islands of observation.

## Summary Table

@tbl-summary below reports the **mean bill length**, **mean flipper
length**, and sample size for each combination of species and sex. Grouping
by both variables shows that, within every species, males tend to have
slightly longer bills and flippers than females -- a pattern worth keeping
in mind before interpreting the scatter plot in the next section.

```{r}
#| label: tbl-summary
#| tbl-cap: "Mean bill length and flipper length (mm) by species and sex, with sample sizes."

penguin_summary <- penguins_clean |>
  group_by(species, sex) |>
  summarise(
    n = n(),
    mean_bill_length_mm = round(mean(bill_length_mm), 1),
    mean_flipper_length_mm = round(mean(flipper_length_mm), 1),
    .groups = "drop"
  )

# kableExtra turns the plain summary table into a styled HTML table:
# striped rows, a bolded header, and hover-highlighting make it easy
# to read on screen.
penguin_summary |>
  kbl(
    col.names = c("Species", "Sex", "n", "Mean Bill Length (mm)",
                  "Mean Flipper Length (mm)"),
    align = "c"
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) |>
  row_spec(0, bold = TRUE, color = "white", background = "#0072B2")
```

## Visualizing the Relationship

@fig-scatter shows bill length plotted against flipper length for every
penguin in the cleaned dataset, colored by species. A few design choices
were made deliberately, following best practices covered in lecture:

1. **Color-blind-safe palette.** The three species are distinguished using
   the Okabe-Ito palette (`#0072B2`, `#D55E00`, `#009E73`), which remains
   distinguishable for the most common forms of color vision deficiency.
2. **Linear trend lines per species**, added with `geom_smooth()`, make the
   *within-species* relationship easy to see even though the species
   overlap in places.
3. **Clear, fully-labeled axes and a legend title**, so the figure can
   stand alone without needing the surrounding text.

```{r}
#| label: fig-scatter
#| fig-cap: "Bill length increases with flipper length within each penguin species. Points are individual penguins; lines are species-specific linear fits with 95% confidence ribbons."
#| fig-width: 7
#| fig-height: 5

ggplot(penguins_clean, aes(x = flipper_length_mm, y = bill_length_mm,
                            color = species)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_color_manual(values = c(
    "Adelie" = "#0072B2",
    "Chinstrap" = "#D55E00",
    "Gentoo" = "#009E73"
  )) +
  labs(
    title = "Bill Length vs. Flipper Length in Palmer Penguins",
    subtitle = "Each species shows a positive within-species relationship",
    x = "Flipper Length (mm)",
    y = "Bill Length (mm)",
    color = "Species"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11, color = "grey30"),
    legend.position = "bottom"
  )
```

Gentoo penguins (green) separate out clearly as the largest-bodied
species, with the longest flippers and comparatively long bills, while
Adelie and Chinstrap penguins overlap more in flipper length but remain
distinguishable by bill length.

## Conclusion

Both the table and the figure tell a consistent story: body size in this
dataset is *not* one-dimensional. Flipper length and bill length both
increase together within each species, but the three species occupy
different regions of that relationship -- meaning either measurement
alone would be enough to make a reasonable guess at species identity for
most individuals. The full, reproducible workflow behind both outputs
(data cleaning, summarizing, and plotting) is documented in the code
chunks above.
