---
title: "Mapping Global Meteorite Falls"
subtitle: "Where meteorites have been seen falling or found, 860-2013"
author: "Yang An"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-location: left
    code-fold: show
    fig-align: center
    embed-resources: true
knitr:
  opts_chunk:
    fig.path: "../output/"
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| echo: false

# tidyverse: wrangling + plotting
# maps: supplies the world polygon data used by map_data("world")
library(tidyverse)
library(maps)
```

## Introduction

This map uses the **Meteorite Landings** dataset (NASA's Meteoritical
Bulletin Database, shared via the
[TidyTuesday](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-06-11)
project), which records over 45,000 meteorites recovered or observed
falling around the world. Each record includes where and when the
meteorite landed, its mass, and whether it was:

- **Fell** -- actually seen falling from the sky, or
- **Found** -- discovered on the ground sometime after landing, with
  no witnessed fall.

The goal here is a single, clean map that shows *where* these events
happened, distinguishes the two record types, and uses point size to
show *how large* each meteorite was.

## Data Preparation

```{r}
#| label: clean-data

meteorites <- read_csv(here::here("Week_07", "data", "meteorites.csv"),
                        show_col_types = FALSE)

# Clean out records that would break or mislead the map:
# - missing coordinates
# - "null island" (0, 0) -- a classic placeholder for missing GPS data,
#   not an actual location in the Gulf of Guinea for 6,000+ meteorites
# - a longitude just over 180 degrees and a year of "2101" are both
#   single obvious data-entry errors, not real locations/dates
# - missing mass, since point size is mapped to mass below
meteorites_clean <- meteorites |>
  drop_na(lat, long, mass) |>
  filter(
    !(lat == 0 & long == 0),
    long >= -180, long <= 180,
    year <= 2016
  )
```

After cleaning, `r nrow(meteorites_clean)` of the original
`r nrow(meteorites)` records remain (the rest were missing coordinates,
placeholder "null island" coordinates, or a couple of obvious
data-entry errors).

## The Map

```{r}
#| label: fig-meteorite-map
#| fig-cap: "Global meteorite landings, colored by whether the fall was witnessed (Fell) or the meteorite was discovered afterward (Found). Point size reflects mass on a log scale."
#| fig-width: 10
#| fig-height: 5.5

world_map <- map_data("world")

ggplot() +
  # Base world map, drawn first so points layer on top of it.
  geom_polygon(
    data = world_map,
    aes(x = long, y = lat, group = group),
    fill = "grey90", color = "white", linewidth = 0.1
  ) +
  # Each meteorite as a point; size shows mass (log scale, since mass
  # ranges from under a gram to tens of thousands of kilograms), color
  # shows whether the fall was witnessed.
  geom_point(
    data = meteorites_clean,
    aes(x = long, y = lat, size = mass, color = fall),
    alpha = 0.25, stroke = 0
  ) +
  # Colorblind-safe two-color palette (Okabe-Ito).
  scale_color_manual(
    values = c(Fell = "#0072B2", Found = "#D55E00"),
    name = "Record type"
  ) +
  # A deliberately narrow size range: with >30,000 overlapping points,
  # large point sizes just merge into solid blobs and hide the
  # continents underneath. Small, mostly-transparent points let
  # geographic density patterns remain visible.
  scale_size_continuous(
    name = "Mass (g, log scale)",
    trans = "log10",
    range = c(0.15, 3),
    labels = scales::label_comma()
  ) +
  coord_quickmap() +
  labs(
    title = "Global Meteorite Landings",
    subtitle = paste0(scales::comma(nrow(meteorites_clean)),
                       " recorded meteorites, 860-2013"),
    caption = "Data: NASA Meteoritical Bulletin via TidyTuesday (2019-06-11)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11, color = "grey30"),
    legend.position = "right"
  )
```

## Interpretation

A few patterns jump out immediately:

1. **"Found" meteorites (orange) are heavily concentrated** in a handful
   of regions -- especially Antarctica and the Sahara -- where dark
   meteorites are easy to spot against ice or sand, and dedicated search
   expeditions have combed the ground for them. This is a *detection*
   pattern as much as a *landing* pattern: meteorites fall roughly
   uniformly over Earth's surface, but we only find a tiny, biased
   sample of them.
2. **"Fell" meteorites (blue) are more evenly spread** across populated
   continents, because they require a human witness -- so they track
   population density rather than any real geographic pattern in where
   meteorites land.
3. **A few very large points** stand out, representing historically
   massive meteorite falls recorded in the database.

The map intentionally keeps latitude/longitude gridlines and axis text
off, since they add clutter without adding information once the
continents themselves are visible -- the title, legend, and caption
carry all the context a reader needs.
